Intrusion Detection with Genetic Algorithms and Fuzzy Logic

### 2013-10-18 paper.tex
Changed what's in some of the tables.

The KDD99 data set is a benchmark data set that was simulated in a military network environment in 1999~\cite{6559603}. It is composed of 41 \emph{features}, which are things that help to distinguish normal connections from attacks.

@@

The detection rate of denial of service attacks in training was 91.64\% and the detection rate of probe attacks in training was 94.79\%. The detection rate of the testing data set increased to 97.92\%. The false positive rate was 1.13\% and the false negative rate was 4.10\%. Detection of an attack happened 2-3 seconds after the packet data arrived at the intrusion detection system. Results from this experiment are shown in Table~\ref{tab:fuzGenExp1}.

@@

In the second experiment, seven tests were run. For each test case there were 13 attack types plus normal activity that were in the training data set. Three attack types were used for the unknown testing data set. For example, the first test case used the training data set that does not have Advance Port Scan, Ack Scan, and Xmas Tree, (which are all probe attacks). These three attacks were then used for the testing data set. Table~\ref{tab:fuzGenExp2} shows some of the results from the fuzzy genetic algorithm and a decision tree algorithm. The fuzzy genetic algorithm has an accuracy of at least 95\% in identifying unknown attacks. When compared with the decision tree algorithm, the fuzzy genetic algorithm has a higher detection rate in all cases except 3 and 5. It can be seen that in cases 2 and 4 the decision tree has low detection rates, while the fuzzy genetic algorithm has much higher detection rates.

\begin{table}
\caption{Unknown Attack Experiment}
\begin{tabular}{|lllcrl|} \hline
Test & Unknown & Decision & & Fuzzy & \\
Case & Attacks & Tree DR & & Genetic DR & \\ \hline

1 & Advance Port Scan & 99.7 & Avg = & 100 & Avg =\\
  & Ack Scan          & 100  & 98.33 & 100 & 100\\
  & Xmas Tree         & 95.3 &       & 100 &\\ \hline

2 & UDP Flood & \textbf{26.7} & Avg = & \textbf{100} & Avg =\\
  & Host Scan & 98.15         & 46.65 & 99.20        & 99.80\\
  & UDP Scan  & \textbf{17.6} &       & \textbf{100} &\\ \hline

3 & Jping    & 100 & Avg =          & 99.20 & Avg =\\
  & Syn Scan & 99.1& \textbf{99.70} & 96.60 & 98.75\\
  & Fin Scan & 100 &                & 100 &\\ \hline

4 & UDP Flood & \textbf{26.6} & Avg = & \textbf{100} & Avg =\\
  & RCP Scan  & 91.56         & 70.35 & 92.60        & 98.15\\
  & Fin Scan  & 99.8          &       & 100          &\\ \hline

5 & Http Flood & 85.7& Avg =          & 100 & Avg =\\
  & RCP Scan  & 92.3 & \textbf{99.94} & 91.80 & 97.50\\
  & Fin Scan  & 99.8 &                & 98.20 &\\
\hline\end{tabular}
\label{tab:fuzGenExp2}
\end{table}

@@

\begin{table}
\caption{Number of Records for KDD99}
\begin{tabular}{|lll|} \hline
Attack & Training & Testing\\
Type   & Data set & Data set\\ \hline
Normal & 39,387 & 39,337\\
DoS & 158,597 & 158,503\\
Probe & 1,550 & 1,674\\
 & &\\
 Total & 199,534 & 199,514\\
\hline\end{tabular}
\label{tab:RecordsForKDD99}
\end{table}

\begin{table}
\caption{Number of Records for RLD09}
\begin{tabular}{|lll|} \hline
Attack & Training & Testing\\
Type   & Data set & Data set\\ \hline
Normal & 8,000 & 16,000\\
DoS & 2,400 & 4,000\\
Probe & 3,900 & 6,500\\
 & &\\
 Total & 14,300 & 26,500\\
\hline\end{tabular}
\label{tab:RecordsForRLD09}
\end{table}

The authors of~\cite{6496342, 6559603} first trained and tested the fuzzy genetic algorithm with the KDD99 data set. There were 6 different types of denial of service attacks and 4 different types of probe attacks. The detection rate of the KDD99 data set was 98.72\%, and it had a false negative rate of 1.55\%. Then 26,500 records of the RLD09 data set were used as the training set. The detection rate was 97.97\%. The results of this experiment are shown in Table~\ref{tab:bothSetsResults}.

@@

\begin{table}
\caption{Results for KDD99 with Certain Attacks}
\begin{tabular}{|cccccc|} \hline
Test & Attack & Type & FP & FN & DR\\
Case & Name   &&&&\\ \hline
1 & Back & DoS & 85.33 & 0.00 & 16.56\\
2 & Land & DoS & 0.39 & 0.00 & 99.61\\
3 & Smurf & DoS & 0.76 & 0.10 & 99.73\\
4 & Portsweep & Probe & 6.40 & 0.00 & 93.66\\
5 & Satan & Probe & 0.74 & 3.75 & 99.22\\
\hline\end{tabular}
\label{tab:kddAttacks}
\end{table}

The final experiment that was run used only the RLD09 data set with the fuzzy genetic algorithm. 17 tests were run, and Table~\ref{tab:rldAttacks} shows the accuracy of detecting some of the cases. The results showed that the detection rate of a majority of the cases were greater than 97.5\%. Again, there were only two cases that had low detection rates, one of which is case 2 in Table~\ref{tab:rldAttacks}.

\begin{table}
\caption{Results for RLD99 with Certain Attacks}
\begin{tabular}{|cccccc|} \hline
Test & Attack & Type & FP & FN & DR\\
Case & Name   &&&&\\ \hline
1 & Smurf & DoS & 0.02 & 0 & 99.98\\
2 & UDP Flood & DoS & 11.06 & 0 & 89.59\\
3 & Ackscan & Probe & 0.03 & 0 & 99.97\\
4 & Ipscan & Probe & 13.01 & 16.4 & 86.89\\
5 & Synscan & Probe & 0.65 & 4.2 & 99.24\\
\hline\end{tabular}
\label{tab:rldAttacks}
\end{table}

>>>>

The KDD99 data set is a benchmark data set that was simulated in a military network environment in 1999~\cite{6559603}. It is composed of 41 \emph{features}, which are things that help to distinguish normal connections from attacks. The experiments that~\cite{6559603, DBLP:journals/corr/abs-1204-1336} ran, which I will describe in Sections~\ref{sec:fuzGenAlgImp} and~\ref{sec:genAlgImp}, used the KDD99 data set.

@@

The detection rate of denial of service attacks in training was 91.64\% and the detection rate of probe attacks in training was 94.79\%. The detection rate of the testing data set increased to 97.92\%. Detection of an attack happened 2-3 seconds after the packet data arrived at the intrusion detection system. Results from this experiment are shown in Table~\ref{tab:fuzGenExp1}.

@@

In the second experiment, seven tests were run. For each test case there were 13 attack types plus normal activity that were in the training data set. Three attack types were used for the unknown testing data set. For example, test case 1 used the training data set that does not have UDP Flood, Host Scan, and UDP Scan, (the first is DoS, and the last two are probe). These three attacks were then used for the testing data set. Table~\ref{tab:fuzGenExp2} shows some of the results from the fuzzy genetic algorithm and a decision tree algorithm. The fuzzy genetic algorithm has an accuracy of at least 95\% in identifying unknown attacks. When compared with the decision tree algorithm, the fuzzy genetic algorithm has a higher detection rate in all cases except 2 and 4. It can be seen that in cases 1 and 3 the decision tree has low detection rates, while the fuzzy genetic algorithm has much higher detection rates.

\begin{table}
\caption{Unknown Attack Experiment}
\begin{tabular}{|llcc|} \hline
Test & Unknown & Decision & Fuzzy\\
Case & Attacks & Tree DR  & Genetic DR\\ \hline

1 & UDP Flood & Avg = & Avg =\\
  & Host Scan & 46.65 & 99.80\\
  & UDP Scan  &       &\\ \hline

2 & Jping    & Avg =          & Avg =\\
  & Syn Scan & 99.70 & 98.75\\
  & Fin Scan &                &\\ \hline

3 & UDP Flood & Avg = & Avg =\\
  & RCP Scan  & 70.35 & 98.15\\
  & Fin Scan  &       &\\ \hline

4 & Http Flood & Avg =          & Avg =\\
  & RCP Scan  & 99.94 & 97.50\\
  & Fin Scan  &                &\\
\hline\end{tabular}
\label{tab:fuzGenExp2}
\end{table}

@@

\begin{table}
\caption{Number of Records for KDD99}
\begin{tabular}{|lll|} \hline
Attack & Training & Testing\\ \hline
Normal & 39,387 & 39,337\\
DoS & 158,597 & 158,503\\
Probe & 1,550 & 1,674\\
 & &\\
 Total & 199,534 & 199,514\\
\hline\end{tabular}
\label{tab:RecordsForKDD99}
\end{table}

\begin{table}
\caption{Number of Records for RLD09}
\begin{tabular}{|lll|} \hline
Attack & Training & Testing\\ \hline
Normal & 8,000 & 16,000\\
DoS & 2,400 & 4,000\\
Probe & 3,900 & 6,500\\
 & &\\
 Total & 14,300 & 26,500\\
\hline\end{tabular}
\label{tab:RecordsForRLD09}
\end{table}

The authors of~\cite{6496342, 6559603} first trained and tested the fuzzy genetic algorithm with the KDD99 data set. There were 6 different types of denial of service attacks and 4 different types of probe attacks. The detection rate of the KDD99 data set was 98.72\%. Then 26,500 records of the RLD09 data set were used as the training set. The detection rate was 97.97\%. The results of this experiment are shown in Table~\ref{tab:bothSetsResults}.

@@

\begin{table}
\caption{Results for KDD99 with Certain Attacks}
\begin{tabular}{|cccccc|} \hline
Test & Attack & Type & FP & FN & DR\\ \hline
1 & Back & DoS & 85.33 & 0.00 & 16.56\\
2 & Smurf & DoS & 0.76 & 0.10 & 99.73\\
3 & Portsweep & Probe & 6.40 & 0.00 & 93.66\\
4 & Satan & Probe & 0.74 & 3.75 & 99.22\\
\hline\end{tabular}
\label{tab:kddAttacks}
\end{table}

The final experiment that was run used only the RLD09 data set with the fuzzy genetic algorithm. 17 tests were run, and Table~\ref{tab:rldAttacks} shows the accuracy of detecting some of the cases. The results showed that the detection rate of a majority of the cases were greater than 97.5\%. Again, there were only two test cases that had low detection rates, one of which is case 2 in Table~\ref{tab:rldAttacks}.

\begin{table}
\caption{Results for RLD99 with Certain Attacks}
\begin{tabular}{|cccccc|} \hline
Test & Attack & Type & FP & FN & DR\\ \hline
1 & Smurf & DoS & 0.02 & 0 & 99.98\\
2 & UDP Flood & DoS & 11.06 & 0 & 89.59\\
3 & Ackscan & Probe & 0.03 & 0 & 99.97\\
4 & Synscan & Probe & 0.65 & 4.2 & 99.24\\
\hline\end{tabular}
\label{tab:rldAttacks}
\end{table}
